---
title: "Where and When Young Drivers Crash"
subtitle: "Targeting Motor Premium Loadings Across Post-COVID Victoria — a report for RACV motor underwriting"
author:
  - "Prashorst Meas"
  - "Min Khant Aung"
  - "Urvi Joshi"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    fig-width: 9
    fig-height: 4.5
    fig-align: center
    code-fold: false
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

# Load the data (filtered to 2018-2024) -- same source path as the report,
# so this deck is reproducible from the same repo.
crashes_raw <- read_csv(
  "data/victorian_road_crash_data_2018-2024.csv",
  show_col_types = FALSE
)

# Cleaning pipeline (identical to the report's Methodology section):
# drop two artefact categories, recode placeholders to NA, type the variables.
crashes <- crashes_raw |>
  filter(
    SEVERITY != "Non injury accident",
    is.na(DTP_REGION) | DTP_REGION != "METROPOLITAN SOUTH EAST REGION"
  ) |>
  mutate(
    hour         = as.integer(sub(":.*", "", ACCIDENT_TIME)),
    SPEED_ZONE   = SPEED_ZONE |> as.character() |> na_if("Not known") |> as.integer(),
    young_driver = YOUNG_DRIVER_18_25 > 0,
    SEVERITY = factor(
      SEVERITY,
      levels = c("Fatal accident", "Serious injury accident", "Other injury accident"),
      labels = c("Fatal", "Serious injury", "Other injury"),
      ordered = TRUE
    )
  )

# Sanity check (matches the report)
stopifnot(nrow(crashes) == 99325L)

# Values used inline on the slides
n_crashes <- nrow(crashes)
yd_rate   <- round(mean(crashes$young_driver) * 100, 1)
```

## The problem

-   Road crashes are a major safety **and financial** concern across Victoria.
-   Young drivers (18–25) are treated as a higher-risk group — but age alone is a blunt instrument.
-   COVID-19 shifted travel patterns, so **where** and **when** crashes happen may have changed.

::: fragment
**Question:** How do crash severity and young-driver involvement vary by **time of day** and **Local Government Area (LGA)** in post-COVID Victoria — and what does that mean for RACV premium loading?
:::

## The dataset

-   **Source:** Victorian Road Crash Data, Dept. of Transport & Planning [@viccrashdata].
-   **Period:** 1 Jan 2018 – 31 Dec 2024.
-   After cleaning: **`r format(n_crashes, big.mark = ",")` crashes** analysed.
-   Key fields: crash severity, hour of day, LGA, region, and young-driver involvement.

```{r}
#| label: tbl-data-summary
#| tbl-cap: "Key variables used in the analysis."

tibble::tribble(
  ~Variable, ~`Role in analysis`,
  "SEVERITY",     "Fatal / Serious injury / Other injury outcome category.",
  "hour",         "Derived from accident time; time-of-day patterns.",
  "LGA_NAME",     "Geographic variation across Local Government Areas.",
  "young_driver", "Whether a crash involved a driver aged 18–25."
) |>
  kable()
```

## Methods

-   **Cleaning:** removed non-injury records and the discontinued *Metropolitan South East Region*; recoded unknown speed zones to missing.
-   **Derived variables:** crash `hour`, ordered `SEVERITY` factor, binary `young_driver` flag (≥1 driver aged 18–25).
-   **Analysis:** descriptive comparison of crash frequency by hour and severity, and young-driver involvement **rates** by LGA (LGAs with ≥100 crashes).
-   Fully reproducible in Quarto — every figure on the following slides is regenerated from the source CSV.

## Results — crash severity composition

```{r}
#| label: fig-severity-overview
#| fig-cap: "Crash severity composition, 2018–2024."

crashes |>
  count(SEVERITY) |>
  ggplot(aes(x = SEVERITY, y = n)) +
  geom_col(fill = "steelblue") +
  labs(x = "Crash severity", y = "Number of crashes") +
  theme_minimal(base_size = 16)
```

-   *Other injury* crashes dominate; fatal crashes are comparatively rare.

## Results — when crashes happen

```{r}
#| label: fig-temporal-patterns
#| fig-cap: "Crash frequency by hour of day and severity, 2018–2024."

crashes |>
  count(hour, SEVERITY) |>
  complete(hour = 0:23, SEVERITY, fill = list(n = 0)) |>
  ggplot(aes(x = hour, y = n, colour = SEVERITY)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) +
  labs(x = "Hour of day", y = "Number of crashes", colour = "Crash severity") +
  theme_minimal(base_size = 16)
```

-   Crashes peak in the **afternoon/early evening (\~3–5 pm)**; lowest 3–5 am.
-   All severity levels follow the same daily rhythm.

## Results — where young drivers crash

```{r}
#| label: fig-young-driver-lga
#| fig-cap: "Top 10 LGAs by young-driver crash involvement rate, 2018–2024."

crashes |>
  filter(!is.na(LGA_NAME)) |>
  group_by(LGA_NAME) |>
  summarise(
    total_crashes        = n(),
    young_driver_crashes = sum(young_driver),
    young_driver_rate    = young_driver_crashes / total_crashes,
    .groups = "drop"
  ) |>
  filter(total_crashes >= 100) |>
  arrange(desc(young_driver_rate)) |>
  slice_head(n = 10) |>
  ggplot(aes(x = reorder(LGA_NAME, young_driver_rate), y = young_driver_rate)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Local Government Area", y = "Young-driver involvement rate") +
  theme_minimal(base_size = 16)
```

-   **Hume** highest, then Corangamite, Casey, Melton (\>30%); spread is moderate, not concentrated.

## Key findings & limitations

-   **Time of day** is a meaningful exposure signal — risk concentrates in peak travel hours.
-   **Geography** adds context: several outer-metro/regional LGAs run higher on young-driver involvement.
-   Variation between LGAs is **moderate**, not extreme — a supporting factor, not a standalone one.
-   **Limitation:** no exposure data (km travelled) or behavioural measures, so no causal claims.

## Conclusions & recommendations

-   Crash risk concentrates in **afternoon/early-evening** hours; young-driver involvement is **unevenly distributed** across LGAs (modest differences).
-   Both **temporal** and **geographic** context add useful information beyond driver age alone.
-   **Recommendation:** incorporate time-of-day and geographic indicators *alongside* age in premium loading — moving toward targeted, evidence-based underwriting.
-   **Future work:** add traffic-exposure and behavioural measures to sharpen risk estimates.

## References

::: {#refs}
:::
