---
title: "Where and When Young Drivers Crash: Targeting Motor Premium Loadings Across Post-COVID Victoria"
subtitle: "A business report for RACV motor underwriting"
author:
  - "Prashorst Meas"
  - "Min Khant Aung"
  - "Urvi Joshi"
date: today
format:
  pdf: default
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

# Methodology and Data Preparation

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# Load the data (filtered to 2018-2024)
crashes_raw <- read_csv(
  "data/victorian_road_crash_data_2018-2024.csv",
  show_col_types = FALSE
)

# Cleaning (documented in the Methodology section)
# Drop two artefact categories, recode placeholders to NA, type the variables.
crashes <- crashes_raw |>
  filter(
    SEVERITY != "Non injury accident",
    is.na(DTP_REGION) | DTP_REGION != "METROPOLITAN SOUTH EAST REGION"
  ) |>
  mutate(
    hour         = as.integer(sub(":.*", "", ACCIDENT_TIME)),
    SPEED_ZONE   = SPEED_ZONE |> as.character() |> na_if("Not known") |> as.integer(),
    young_driver = YOUNG_DRIVER_18_25 > 0,
    SEVERITY = factor(
      SEVERITY,
      levels = c("Fatal accident", "Serious injury accident", "Other injury accident"),
      labels = c("Fatal", "Serious injury", "Other injury"),
      ordered = TRUE
    )
  )

# Sanity check for errors
stopifnot(nrow(crashes) == 99325L)

# Values used inline
n_crashes <- nrow(crashes)
yd_rate   <- round(mean(crashes$young_driver) * 100, 1)
```

\<-- Methodology part here (300 words) --\> Also need to include figure or table as assignment requirement

# Results

\<-- results-temporal part here –\>

## Geographic results for young-driver crash analysis

```{r}
echo: false

#| label: fig-young-driver-lga
#| fig-cap: "Top 10 Victorian LGAs by young-driver crash involvement rate, 2018–2024."

young_driver_lga <- crashes |>
  filter(!is.na(LGA_NAME)) |>
  group_by(LGA_NAME) |>
  summarise(
    total_crashes = n(),
    young_driver_crashes = sum(young_driver),
    young_driver_rate = young_driver_crashes / total_crashes,
    .groups = "drop"
  ) |>
  filter(total_crashes >= 100) |>
  arrange(desc(young_driver_rate)) |>
  slice_head(n = 10)

ggplot(
  young_driver_lga,
  aes(
    x = reorder(LGA_NAME, young_driver_rate),
    y = young_driver_rate
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  labs(
    x = "Local Government Area (LGA)",
    y = "Young-driver crash involvement rate"
  ) +
  theme_minimal()

```

As shown in \@fig-young-driver-lga, young-driver crash involvement varied across Victorian Local Government Areas (LGAs) between 2018 and 2024. Hume recorded the highest proportion of crashes involving drivers aged 18–25, followed by Corangamite, Casey, and Melton, with young-driver involvement rates exceeding 30% of total crashes. Several outer metropolitan LGAs, including Casey, Whittlesea, and Wyndham, also appeared among the highest-ranked areas, suggesting elevated exposure or behavioural risk among younger drivers in growth corridors. However, differences between LGAs were relatively modest, indicating that elevated young-driver involvement is a widespread issue rather than being isolated to only a few regions. From an underwriting perspective, these findings suggest that geographic variation may provide useful contextual information for premium loading decisions, particularly when combined with driver age and time-of-day risk indicators.
