---
title: "Where and When Young Drivers Crash: Targeting Motor Premium Loadings Across Post-COVID Victoria"
subtitle: "A business report for RACV motor underwriting"
author:
  - "Prashorst Meas"
  - "Min Khant Aung"
  - "Urvi Joshi"
date: today
format:
  pdf:
    toc: true
    number-sections: true
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

# Executive Summary

\<--After all the parts are finished, write here--\>

# Introduction

Road crashes remain a significant public safety and financial concern across Victoria, contributing to injury, loss of life, and substantial insurance costs. Young drivers are often considered a higher-risk group due to limited driving experience and greater exposure to risky driving behaviours. Following the COVID-19 pandemic, traffic patterns and travel behaviours shifted considerably, potentially changing the spatial and temporal distribution of crash risk across the state. Understanding where and when crashes involving young drivers are more likely to occur can provide valuable insight for risk assessment and insurance decision-making.

This report investigates how crash severity and young-driver involvement vary by time of day and Local Government Area (LGA) across post-COVID Victoria using Victorian Road Crash Data from 2018 to 2024. The analysis is framed from the perspective of an RACV motor underwriting analyst seeking to identify geographic and temporal risk factors relevant to premium loading decisions. Particular attention is given to young-driver involvement, crash severity, and regional variation to identify areas of elevated risk. Findings from this analysis may assist insurers in developing more targeted and evidence-based underwriting strategies.

# Methodology and Data Preparation

```{r}
#| label: setup
#| warning: false
#| message: false
#| echo: false
library(tidyverse)
library(knitr)

# Load the data (filtered to 2018-2024)
crashes_raw <- read_csv(
  "data/victorian_road_crash_data_2018-2024.csv",
  show_col_types = FALSE
)

# Cleaning (documented in the Methodology section)
# Drop two artefact categories, recode placeholders to NA, type the variables.
crashes <- crashes_raw |>
  filter(
    SEVERITY != "Non injury accident",
    is.na(DTP_REGION) | DTP_REGION != "METROPOLITAN SOUTH EAST REGION"
  ) |>
  mutate(
    hour         = as.integer(sub(":.*", "", ACCIDENT_TIME)),
    SPEED_ZONE   = SPEED_ZONE |> as.character() |> na_if("Not known") |> as.integer(),
    young_driver = YOUNG_DRIVER_18_25 > 0,
    SEVERITY = factor(
      SEVERITY,
      levels = c("Fatal accident", "Serious injury accident", "Other injury accident"),
      labels = c("Fatal", "Serious injury", "Other injury"),
      ordered = TRUE
    )
  )

# Sanity check for errors
stopifnot(nrow(crashes) == 99325L)

# Values used inline
n_crashes <- nrow(crashes)
yd_rate   <- round(mean(crashes$young_driver) * 100, 1)
```

## Methodology

```{r}
#| echo: false
#| label: fig-severity-overview
#| fig-cap: "Crash severity composition in the cleaned Victorian road crash dataset, 2018–2024."

crashes |>
  count(SEVERITY) |>
  ggplot(aes(x = SEVERITY, y = n)) +
  geom_col() +
  labs(
    x = "Crash severity",
    y = "Number of crashes"
  ) +
  theme_minimal()
```

```{r}
#| echo: false
#| label: tbl-data-summary
#| tbl-cap: "Summary of key variables used in the analysis."

key_variable_summary <- tibble::tribble(
  ~Variable, ~Role_in_analysis,
  "SEVERITY", "Crash outcome category used to compare fatal, serious injury, and other injury crashes.",
  "hour", "Derived from accident time and used to analyse time-of-day crash patterns.",
  "LGA_NAME", "Used to compare geographic variation across Local Government Areas.",
  "DTP_REGION", "Used for broad regional comparison across Victoria.",
  "young_driver", "Binary indicator showing whether a crash involved at least one driver aged 18–25."
)

knitr::kable(key_variable_summary)
```

Victorian Road Crash Data was obtained from the Victorian Department of Transport and Planning and filtered to include crashes occurring between 1 January 2018 and 31 December 2024. The dataset contains information on crash severity, location, road conditions, vehicle involvement, and demographic indicators relevant to motor insurance risk assessment. Data preparation involved removing non-injury accident records and excluding entries associated with the discontinued “Metropolitan South East Region” category to improve consistency across regional comparisons. Placeholder values, such as unknown speed zones, were recoded as missing values, while additional analytical variables were constructed, including crash hour and a binary indicator of young-driver involvement, defined as crashes involving at least one driver aged 18–25. Crash severity was also recoded into an ordered factor comprising Fatal, Serious injury, and Other injury categories.

As shown in @fig-severity-overview, the cleaned dataset was first examined according to crash severity composition to understand the crash outcomes included in the analysis. Table @tbl-data-summary summarises the key variables used to investigate temporal and geographic crash patterns. The Results section then examines time-of-day crash patterns and Local Government Area variation in young-driver crash involvement to support RACV motor underwriting decisions.

# Results

## Temporal Crash Patterns

```{r}
#| echo: false
#| label: fig-temporal-patterns
#| fig-cap: "Crash frequency by hour of day and severity in Victoria, 2018–2024."

hourly_crashes <- crashes |>
  count(hour, SEVERITY) |>
  complete(
    hour = 0:23,
    SEVERITY,
    fill = list(n = 0)
  )

ggplot(
  hourly_crashes,
  aes(x = hour, y = n, colour = SEVERITY)
) +
  geom_line(linewidth = 1) +
  scale_x_continuous(
    breaks = seq(0, 23, by = 2)
  ) +
  labs(
    x = "Hour of day",
    y = "Number of crashes",
    colour = "Crash severity"
  ) +
  theme_minimal()
```

As shown in \@fig-temporal-patterns, crash frequency varied considerably across the day between 2018 and 2024. Crashes involving all severity levels generally increased during morning commuting hours, followed by a more pronounced rise during the afternoon and early evening period. The highest concentration of crashes occurred between approximately 3:00 pm and 5:00 pm, while crash frequency was substantially lower during early morning hours between 3:00 am and 5:00 am.

Other injury crashes consistently represented the largest category across all hours, while serious injury and fatal crashes remained comparatively less frequent. However, serious and fatal crashes still followed a similar temporal pattern, with increased activity during peak travel periods. These findings suggest that time of day may be an important contextual factor for underwriting decisions, particularly during periods of higher traffic exposure when crash frequency increases substantially.

## Geographic results for young-driver crash analysis

```{r}
#| echo: false

#| label: fig-young-driver-lga
#| fig-cap: "Top 10 Victorian LGAs by young-driver crash involvement rate, 2018–2024."

young_driver_lga <- crashes |>
  filter(!is.na(LGA_NAME)) |>
  group_by(LGA_NAME) |>
  summarise(
    total_crashes = n(),
    young_driver_crashes = sum(young_driver),
    young_driver_rate = young_driver_crashes / total_crashes,
    .groups = "drop"
  ) |>
  filter(total_crashes >= 100) |>
  arrange(desc(young_driver_rate)) |>
  slice_head(n = 10)

ggplot(
  young_driver_lga,
  aes(
    x = reorder(LGA_NAME, young_driver_rate),
    y = young_driver_rate
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  labs(
    x = "Local Government Area (LGA)",
    y = "Young-driver crash involvement rate"
  ) +
  theme_minimal()

```

As shown in \@fig-young-driver-lga, young-driver crash involvement varied across Victorian Local Government Areas (LGAs) between 2018 and 2024. Hume recorded the highest proportion of crashes involving drivers aged 18–25, followed by Corangamite, Casey, and Melton, with young-driver involvement rates exceeding 30% of total crashes. Several outer metropolitan LGAs, including Casey, Whittlesea, and Wyndham, also appeared among the highest-ranked areas, indicating geographic variation in crashes involving young drivers. However, differences between LGAs were relatively modest, suggesting that elevated young-driver involvement occurs across multiple regions rather than being isolated to a small number of areas. From an underwriting perspective, these findings indicate that geographic context may provide useful supporting information for premium loading decisions when considered alongside driver age and temporal crash risk indicators.
